---
title: "An Introduction to celda"
author: "Josh D. Campbell, Sean Corbett, Masanao Yajima, Zichun Liu, Shiyi Yang, Tianwen Huan, Anahita Bahri"
date: "`r Sys.Date()`"
output: pdf_document
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,fig.width=8,fig.height=7 ,
			out.width="0.7\\linewidth", fig.align = "center")
library(RColorBrewer)
library(ggplot2)
library(celda)
library(gtools)
library(pheatmap)
library(reshape2)
library(vcd)
library(corrplot)
```

## 1 Philosophy

The purpose of this vignette is to briefly introduce package `celda`. `celda` is the abbreviation of "CELlular Latent Dirichlet Allocation."

It means that we implement Bayesian Hierarchical Modeling for Clustering Single-Cell RNA-Seq Data

To install `celda`:  

```{r,eval=FALSE}
library(devtools)
install_github("definitelysean/celda")
```
  
## 2 Single-Cell RNA-Seq Data

Single Cell RNA-Seq Data

## 3 Latent Dirichlet allocation

In natural language processing, latent Dirichlet allocation (LDA) is a generative statistical model that allows sets of observations to be explained by unobserved groups that explain why some parts of the data are similar.  

For example, if observations are words collected into documents, it posits that each document is a mixture of a small number of topics and that each word's creation is attributable to one of the document's topics.  

LDA is an example of a topic model and was first presented as a graphical model for topic discovery by David Blei, Andrew Ng, and Michael I. Jordan in 2003. Essentially the same model was also proposed independently by J. K. Pritchard, M. Stephens, and P. Donnelly in the study of population genetics in 2000. Both papers have been highly influential, with 16488 and 18170 citations respectively by December 2016.([via](https://en.wikipedia.org/wiki/Latent_Dirichlet_allocation))

## 4 An example using simulation data
With the cursory background out of the way, we’ll proceed with some examples to help understand how the package is used.

### 4.1 Generate data using built-in function

celda provide a built-in method to generate simulation data. Directly call `cCG.generateCells`. We can specify some arguments in the function:

S:
 
C.Range: 
 
N.Range:
 
G:
 
K: Cell number
 
L: Gene number
 
alpha:
 
beta:
 
gamma:
 
delta:
 
seed:

```{r simulate data}
sim_counts = cCG.generateCells(G = 100,K = 3, L = 5)
str(sim_counts)
```


The `cCG.generateCells` will return a list with standard format of celda. In the output:
  
z:
 
y:
 
sample: cell.sample.label
 
counts: cell.counts
 
theta:
 
phi:
 
psi: 
 
eta:



### 4.2 Filter cells with high gene expressions

Continue using simulation data. We now perform some filtering in order to obtain obvious result.

```{r filter high expression}
ind<-rowSums(sim_counts$counts>4)>4 #filter cells with high gene expression
counts = sim_counts$counts[ind,]
dim(counts)
```

### 4.3 LDA Clustering on Cells and Genes

We cluster cells and genes using `celda_CG`, here we want to cluster cells into 3 categories and cluster genes into 5 categories. And due to time consumption, we limit the iteration to be 25.

K = Total number of cell populations
L = Total number of transcriptional states

```{r}
celda_cluster_assignments = celda_CG(counts, sample.label = sim_counts$sample,
                                     K = 3, L = 5, max.iter = 25)
str(celda_cluster_assignments)
```

z: Cell cluster assignments from the final iteration of Gibbs sampling
 
y: gene cluster assignments from the final iteration of Gibbs sampling 

complete.z: History of cell cluster assignments across all iterations of Gibbs

complete.y: History of gene cluster assignments across all iterations of Gibbs

z.stability: [0,1] measure of stability for the cell clustering chain

y.stability: [0,1] measure of stability for the gene clustering chain

complete.z.stability: History of z.stability over all iterations of Gibbs sampling

complete.y.stability: Historyof y.stability over all iterations of Gibbs sampling

completeLogLik: Log-likelihood of all gene and cell cluster assignments over all iterations of Gibbs sampling

finalLogLik: Log-likelihood of final gene and cell cluster assignments

z.prob:

y.prob:

### 4.4 Visualization

#### 4.4.1 Heatmap, Mosaic plot and corrplot on Cluster cell data
```{r }
samp = sim_counts$sample

cluster_z = celda_cluster_assignments$z

tb = xtabs(~samp + cluster_z)

nmat <- tb/rowSums(tb)

mnmat<-melt(nmat)

mnmat$samp<-factor(mnmat$samp)

mnmat$cluster_z<-factor(mnmat$cluster_z,levels=sort(unique(mnmat$cluster_z),TRUE))
```

```{r fig.width=5,fig.height=3 }
ggplot(mnmat)+geom_tile()+
  aes(x=samp,y=cluster_z,fill=value)+scale_fill_gradient2()+
  scale_x_discrete(position = "top")+xlab("sample")+ylab("cluster")

mosaic(tb)

corrplot(t(nmat),method = "circle",is.corr = FALSE)
```


#### 4.4.2 Use celda_heatmap to produce heatmap

```{r}
celda_heatmap(counts,
              K=3,z = celda_cluster_assignments$z,
              L = 10,y = celda_cluster_assignments$y)

celda_heatmap(log(counts+1),
              K=3,z = celda_cluster_assignments$z, 
              L = 10,y = celda_cluster_assignments$y)

celda_heatmap(t(apply(counts,1,scale)),
              K=3,z = celda_cluster_assignments$z, 
              L = 10,y = celda_cluster_assignments$y)
```

### 4.5 MCMC Diagnostic
```{r}
```


## 5 An example using simulation data

### 5.1 Read in data

To be filled 

```{r}
#corresponding read file and compile code.
```

