---
title: "DecontX has moved to the decontX package"
author:
- name: Shiyi (Iris) Yang
  affiliation: &id Boston University School of Medicine
- name: Zhe Wang
  affiliation: *id
- name: Yuan Yin
  affiliation: *id
- name: Joshua Campbell
  affiliation: *id
  email: camp@bu.edu
date: "`r Sys.Date()`"
output:
  BiocStyle::html_document:
    toc: true
vignette: >
  %\VignetteIndexEntry{Estimate and remove cross-contamination from ambient RNA in single-cell data with DecontX}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# DecontX now lives in the decontX package

DecontX is a Bayesian method that estimates and removes contamination from
ambient RNA in individual cells of droplet-based single-cell RNA-seq data.
Since celda 1.29.1, DecontX is developed and documented in the standalone
[decontX](https://bioconductor.org/packages/decontX/) Bioconductor package.

celda's `decontX()`, `decontXcounts()`, `plotDecontXContamination()`,
`plotDecontXMarkerExpression()`, `plotDecontXMarkerPercentage()`, and
`simulateContamination()` are thin wrappers that pass their arguments on to
the decontX package. Code that calls them through celda keeps working, as
long as decontX is installed. New code can call the decontX package
directly.

# Installation

Install decontX from Bioconductor:

```{r install, eval=FALSE}
if (!requireNamespace("BiocManager", quietly = TRUE)) {
    install.packages("BiocManager")
}
BiocManager::install("decontX")
```

# Tutorial

The full tutorial, which decontaminates the 10X PBMC 4k dataset, is in the
decontX vignette. Read it on the
[decontX Bioconductor page](https://bioconductor.org/packages/decontX/), or
open it in R after installing decontX:

```{r openVignette, eval=FALSE}
vignette("decontX", package = "decontX")
```

The decontX package also has a vignette for DecontPro, which decontaminates
protein expression data from single cells.

To report a bug or ask a question about DecontX, use the decontX GitHub page
at https://github.com/campbio/decontX.

# Session Information

```{r sessionInfo}
sessionInfo()
```
