---
title: "Decontamination of ambient RNA in single-cell genomic data with DecontX" 
author:
- name: Shiyi (Iris) Yang
  affiliation: *id
- name: Zhe Wang
  affiliation: *id
- name: Joshua Campbell
  affiliation: *id
  email: camp@bu.edu
date: "`r Sys.Date()`"
output:
  BiocStyle::html_document:
    toc: true
vignette: >
  %\VignetteIndexEntry{Estimate and remove cross-contamination from ambient RNA in single-cell data with DecontX}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setoptions, echo=FALSE}
library(knitr)
knitr::opts_chunk$set(dev = "png")
```

# Installation

celda can be installed from Bioconductor:

```{r install, eval= FALSE}
if (!requireNamespace("BiocManager", quietly=TRUE)) {
    install.packages("BiocManager")}
BiocManager::install("celda")
```

The package can be loaded using the `library` command.

```{r load, eval=TRUE, message=FALSE}
library(celda)
```

DecontX can take either `SingleCellExperiment` object from package [SingleCellExperiment package](https://bioconductor.org/packages/release/bioc/html/SingleCellExperiment.html) or a single counts matrix as input. `decontX` will attempt to convert any input matrix to class `dgCMatrix` from package [Matrix](https://cran.r-project.org/web/packages/Matrix/index.html) before beginning any analyses.


# Load PBMC4k data from 10X

We will utlize the 10X PBMC 4K dataset as an example. This can be easily retrieved from the package [TENxPBMCData](http://bioconductor.org/packages/release/data/experiment/html/TENxPBMCData.html). Make sure the the column names are set before running decontX.

```{r load_10X, eval=TRUE, message=FALSE}
library(TENxPBMCData)  
pbmc4k <- TENxPBMCData("pbmc4k")
colnames(pbmc4k) <- paste(pbmc4k$Sample, pbmc4k$Barcode, sep="_")
rownames(pbmc4k) <- rowData(pbmc4k)$Symbol_TENx
```


# Running decontX

To run decontX with a SingleCellExperiment object, simply use the following command. 

```{r decontX, eval=TRUE, message=FALSE}
pbmc4k <- decontX(x = pbmc4k) 
```

The contamination can be found in the `colData(pbmc4k)$decontX_contamination` and the decontaminated counts can be accessed with `decontXcounts(pbmc4k)`. If the input object was a matrix, make sure to save the output into a variable with a different name (e.g. result). The result object will be a list with contamination in `result$contamination` and the decontaminated counts in `result$decontXcounts`. 

# Plotting DecontX results

## Cluster labels on UMAP
DecontX creates a UMAP which we can use to plot the cluster labels automatically identified in the analysis. Note that the clustering approach used here is designed to find "broad" cell types rather than individual cell subpopulations within a cell type. 

```{r}
umap <- reducedDim(pbmc4k, "decontX_UMAP")
plotDimReduceCluster(umap[,1], umap[,2], cluster = pbmc4k$decontX_clusters)
```
The estimated percentage of contamination in each cell can also be plotted on a UMAP.

## Contamination on UMAP
The percentage of contamination in each cell can be plotting on the UMAP to visualize what what clusters may have higher levels of ambient RNA.

```{r}
plotDecontXContamination(pbmc4k)
```

## Expression of markers on UMAP
Known marker genes can also be plotted on the UMAP to identify the cell types for each cluster. We will use CD3D and CD3E for T-cells, LYZ, S100A8, and S100A9 for monocytes, CD79A, CD79B, and MS4A1 for B-cells, GNLY for NK-cells, and PPBP for megakaryocytes.

```{r}
temp <- as.matrix(counts(pbmc4k))
temp <- temp[rowSums(temp) > 0,]
plotDimReduceFeature(umap[,1], umap[,2], temp, features = c("CD3D", "CD3E", "GNLY", "LYZ", "S100A8", "S100A9", "CD79A", "CD79B", "MS4A1"), exactMatch = TRUE)
```

## Barplot of markers detected in cell clusters
The percetage of cells within a cluster that have detectable expression of marker genes can be displayed in a barplot. Markers for cell types need to be supplied in a named list. First, the detection of marker genes in the original `counts` assay is shown:

```{r barplotCounts}
markers <- list(Tcell_Markers=c("CD3E", "CD3D"),
                    Bcell_Markers=c("CD79A", "CD79B", "MS4A1"), 
                    Monocyte_Markers=c("S100A8", "S100A9", "LYZ"),
                    NKcell_Markers="GNLY")
cellTypeMappings <- list(Tcells=2, Bcells=5, Monocytes=1, NKcells=6)
plotDecontXMarkers(pbmc4k, markers = markers, groupClusters = cellTypeMappings, assayName = "counts")
```

We can then look to see how much DecontX removed aberrant expression of marker genes in each cell type by changing the `assayName` to be `decontXcounts`:

```{r barplotDecontCounts}
plotDecontXMarkers(pbmc4k, markers = markers, groupClusters = cellTypeMappings, assayName = "decontXcounts")
```

Percentages of marker genes detected in other cell types were reduced or completely removed. For example, the percentage of cells that expressed Monocyte marker genes was greatly reduced in T-cells, B-cells, and NK-cells. Some important considerations when making barplots:

* Cell clusters can be renamed and re-grouped using the `groupCluster` parameter, which also needs to be a named list. If `groupCluster` is used, cell clusters not included in the list will be excluded in the barplot. For example, if we wanted to group T-cells and NK-cells together, we could
* The level a gene needs to be expressed to be considered detected in a cell can be adjusted using the `threshold` parameter.
* If you are not using a `SingleCellExperiment`, then you will need to supply the original counts matrix or the decontaminated counts matrix as the first argument to generate the barplots. 


```{r}
sessionInfo()
```

