---
title: "Estimate and remove cross-contamination from ambient RNA for scRNA-seq data with DecontX"
author: "Shiyi Yang, Sean Corbett, Yusuke Koga, Zhe Wang, W. Evan Johnson, Masanao Yajima, Joshua D. Campbell"
date: "`r Sys.Date()`"
output:
  BiocStyle::html_document:
    toc: true
vignette: >
  %\VignetteIndexEntry{Estimate and remove cross-contamination from ambient RNA for scRNA-seq data with DecontX}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---


# Introduction 
DecontX is a Bayesian hierarchical model to estimate and remove cross-contamination from ambient RNA in single-cell RNA-seq count data generated from droplet-based sequencing devices.DecontX will take the count matrix with/without the cell labels and estimate the contamination level and deliver a decontaminted count matrix for downstream analysis. 

In this vignette we will demonstrate how to use DecontX to estimate and remove contamination.  


The package can be loaded using the `library` command.

```{r, eval=TRUE, message=FALSE} 
library(celda)
```
To see the latest updates and releases or to post a bug, see our GitHub page at https://github.com/compbiomed/celda. To ask questions about running Celda, visit our Google group at https://groups.google.com/forum/#!forum/celda-list.


# Generation of a cross-contaminated dataset 
DecontX will take a matrix of counts (referred as observed counts) where each row is a feature, each column is a cell, and each entry in the matrix is the number of counts of each feature in each cell. To illustrate the utility of DecontX, we will apply it to a simulated dataset.

In the function `simulateContaminatedMatrix`, the K parameter designates the number of cell clusters, the C parameter determines the number of cells, the G parameter determines the number of genes in the simulated dataset.

```{r}
sim_counts = simulateContaminatedMatrix( G = 300, C = 100, K = 3 ) 
```

```{r}
contamination = sim_counts$observedCounts - sim_counts$nativeCounts 
```

The `z` variable contains the population label for each cell
```{r}
table(sim_counts$z) 
```

The `phi` and `eta` variables contain the expression distributions and contamination distributions for each population, respectively. Each column corresponds to a population, each row represents a gene. The sum of the rows equal to 1. 
```{r}
colSums( sim_counts$phi ) 
colSums( sim_counts$eta )
```


# Decontamination using DecontX
DecontX uses bayesian method to estimate and remove contamination via varitaional inference. 
```{r, warning = FALSE, message = FALSE}
decontx.model = DecontX( counts = sim_counts$observedCounts, z = sim_counts$z ) 
```

## Check convergance
Use log-likelihood to check convergance 
```{r, eval = TRUE, fig.width = 5, fig.height = 5}
plot( decontx.model$res.list$logLikelihood ) 
```

## Evaluate model performance 
`DecontX` estimates a contamination proportion for each cell. We compare the estimated contamination proportion with the real contamination proportion.
```{r, eval = TRUE, fig.width = 5, fig.height = 5}
plot( decontx.model$res.list$est.conp, colSums(contamination) / sim_counts$N.by.C,  col=sim_counts$z) ; abline( 0, 1) 
```
